---
title: "| RESEARCH PROTOCOL\n| \n| Assessment of bias within existing prediction models\n"
fontsize: 12pt
geometry: margin=1in
output:
  bookdown::html_document2:
    df_print: paged
    toc: yes
    toc_depth: 2
    toc_float: yes
    number_sections: yes
    number_tables: yes
    css: "style.css"
  bookdown::pdf_document2:
    keep_tex: yes
    latex_engine: xelatex
    md_extensions: +raw_attribute
    number_sections: yes
    # citation_package: natbib
    includes:
      before_body: title.tex
  bookdown::word_document2:
    toc: yes
header-includes:
- \usepackage[numbers,sort&compress]{natbib}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{caption}
- \usepackage{rotating}
- \usepackage{multirow}
- \usepackage{mwe,tikz}
- \usepackage[percent]{overpic}
- \usepackage{enumitem}
- \usepackage{hyperref}
- \newcolumntype{P}[1]{>{\raggedright\arraybackslash}p{#1}}
- \newcommand{\footerDate}{`r params$date`}
- \input{header.tex}
longtable: yes
mainfont: Arial
bibliography: Protocol.bib
params:
  date: '2021-03-20'
  version: 1.0.0
subtitle: 'Version: `r params$version`'
link-citations: true
csl: jamia.csl
---
    
    ```{r setup, include=FALSE}
  knitr::opts_chunk$set(echo = TRUE)
  options(kableExtra.latex.load_packages = FALSE)
  library(kableExtra)
  #knitr::knit_hooks$set(document = function(x) {sub('\\usepackage[]{color}', '\\usepackage[table]{xcolor}', x, fixed = TRUE)})
  library(dplyr)
  options(knitr.kable.NA = "")
  if (!knitr::is_latex_output() && !knitr::is_html_output()) {
    options(knitr.table.format = "simple")
  }
  pdf2png <- function(path) {
    # only do the conversion for non-LaTeX output
    if (knitr::is_latex_output()) {
      return(path)
    }
    path2 <- xfun::with_ext(path, "png")
    img <- magick::image_read_pdf(path)
    magick::image_write(img, path2, format = "png")
    path2
  }
  latex_table_font_size <- 8
  #source("PrintCohortDefinitions.R")
  #numberOfNcs <- nrow(readr::read_csv("../inst/settings/NegativeControls.csv", col_types = readr::cols()))
  ```
  
   List of Abbreviations
  
  ```{r abbreviations, echo=FALSE, results="asis", warning=FALSE}
  abbreviations <- readr::read_delim(col_names = FALSE, delim = ";", trim_ws = TRUE, file = "
  AUC; Area Under the receiver-operator Curve 
  CCAE; IBM MarketScan Commercial Claims and Encounters 
  CDM; Common Data Model
  CRAN; Comprehensive R Archive Network
  EHR; Electronic Health Record
  IRB; Institutional review board
  MDCR; IBM MarketScan Medicare Supplemental Database
  MDCD; IBM MarketScan Multi-State Medicaid Database 
  OHDSI; Observational Health Data Science and Informatics
  OMOP; Observational Medical Outcomes Partnership
")
  tab <- kable(abbreviations, col.names = NULL, linesep = "", booktabs = TRUE)
  if (knitr::is_latex_output()) {
    tab %>% kable_styling(latex_options = "striped", font_size = latex_table_font_size)
  } else {
    tab %>% kable_styling(bootstrap_options = "striped")
  } 
  ```
  
  ```{r parties, echo=FALSE, results="asis", warning=FALSE}
  parties <- readr::read_delim(col_names = TRUE, delim = ";", trim_ws = TRUE, file = "
  Investigator; Institution/Affiliation
  Ross Williams *; Erasmus University Medical Center
")
  tab <- kable(parties, booktabs = TRUE, linesep = "") %>% 
    column_spec(1, width = "10em") %>%
    column_spec(2, width = "30em") %>%
    footnote(general = "* Principal Investigator", general_title = "")
  if (knitr::is_latex_output()) {
    tab %>% kable_styling(latex_options = "striped", font_size = latex_table_font_size)
  } else {
    tab %>% kable_styling(bootstrap_options = "striped")
  }
  ```
  
  ## Disclosures
  
  This study is undertaken within Observational Health Data Sciences and Informatics (OHDSI), an open collaboration.
  This study recieved funding from EHDEN an IMI funded project.
  
  # Abstract
  
  **Background and Significance**
    
  Patient-level prediction models in healthcare can help identify a patients personalized risk of some future medical event. These models can aid medical decision making. However, in many areas of prediction study, developed models have been shown to exhibit bias against subgroups<refs>. In the medical field, the topic of bias has been given increasing improtance over the past few years. If models are to be adopted within clinical practice that systematically over or underestimate risk for protected subgroups of patients, this can have a negative and unfair impact on the treatment decisions made for that group. 
  
  **Study Aims**
    
   We aim to answer the question: Are models that are currntly used in clinical practice biased against subgroups of patients?
  
  **Study Description**
    
  * Design: Retrospecive cohort with external validation  
  * Target Populations: 
    1) QCOVID (is uk specific, could be an issue)  Adults aged 19-100 years with one or two doses of covid-19 vaccination between 8 December 2020 and 15 June 2021.
    2) SEEK COVER: Patients with an initial interaction with a healthcare provider with a diagnosis of C19
    3) Chads2Vasc: non-valvular AF
    4) ATRIA Bleeding Risk Score: Patients newly intiating warfarin
    
  * Outcomes: 
    1) Primary outcome was covid-19 related death. (from 14 days post vac)
    2) covid severity
    3) stroke
    4) Hemorrhage

  * Time-at-risk: 
    1) 1 days to unknown days (would set this to a year)
    2) 1 days to 30 days
    3) 1 days to 365 days
    4) 1 days to 365 days
    
  * Covariates
    
    
     QCOVID
     
     
     Seek Cover
      1) Age
      2) Sex
      3) cancer
      4) copd
      5) diabetes
      6) heart disease
      7) hypertension
      8) hyperlipidaemia
      9) kidney disease

    
    Chads2Vasc
      1) Age
      2) Sex
      3) CHF history
      4) Hypertension history
      5) Stroke/TIA/thromboembolism history
      6) Vascular disease history (prior MI, peripheral artery disease, or aortic plaque)
      7) Diabetes history

    
    ATRIA Bleeding
      1) Anemia
      2) Severe Renal Disease
      3) Age >= 75 years
      4) Any Prior Hemorrhage Diagnosis
      5) Hypertension History

    * Models:
    QRisk
    SEEK COVER
    Chads2Vasc
    ATRIA bleeding
    
  * Internal Metrics:
    - Area Under the receiver-operator Curve (AUC).
    - Sensitivity and PPV at a range of thresholds.
    - Calibration Plots  
    - Net Benefit
  * External Metrics:
    - Area Under the receiver-operator Curve (AUC).
    - Sensitivity and PPV at a range of thresholds.
    - Calibration Plots  (pre and post recalibration)
    - Net Benefit (pre and post recalibration)
  * Fairness metrics:
    [ADD]


# Amendments and Updates

```{r amendments, echo=FALSE, results="asis", warning=FALSE}
amendments <- readr::read_delim(col_names = TRUE, delim = ";", trim_ws = TRUE, file = "
  Number; Date; Section of study protocol; Amendment or update; Reason
  None;;;; 
")
tab <- kable(amendments, booktabs = TRUE, linesep = "")
if (knitr::is_latex_output()) {
  tab %>% kable_styling(latex_options = "striped", font_size = latex_table_font_size)
} else {
  tab %>% kable_styling(bootstrap_options = "striped")
}
```

# Milestones

```{r dates, echo=FALSE, results="asis", warning=FALSE}
dates <- readr::read_delim(col_names = TRUE, delim = ";", trim_ws = TRUE, file = "
  Milestone; Planned / actual date
  Start of analysis;
  End of analysis;
  Results presentation;
")
tab <- kable(dates, booktabs = TRUE, linesep = "") 
if (knitr::is_latex_output()) {
  tab %>% kable_styling(latex_options = "striped", font_size = latex_table_font_size)
} else {
  tab %>% kable_styling(bootstrap_options = "striped")
}
```

# Rationale and Background

[ADD]

# Study Objectives

Do existing models exhibit bias when performance is analysed for specific patient subgroups?

Specific aims:
  
[ADD]

# Research Methods

## Target Populations



## Outcomes and Time-at-risk


## Patient sub populations of interest
 1) sex
 2) various Race (this will need to be country specific)
 3) intersection of above?

## Metrics

We will evaluate the performance by calculating the area under the receiver operating characteristic curve for overall discrimination and calibration plots for calibration.  We will also calculate the area under the precision recall curve and net benefit plots.  Threshold dependent performance will be displayed using the probability threshold plot.

We will also implement multiple fairness metrics.

Performance will be assessed by comparing the models overall perfroamcne to the performance within specific subgroups of patients

## Data sources

We will execute the study as an OHDSI network study.
All data partners within OHDSI are encouraged to participate voluntarily and can do so conveniently, because of the community's shared Observational Medical Outcomes Partnership (OMOP) common data model (CDM) and OHDSI tool-stack.

Many OHDSI community data partners have already committed to participate and we will recruit further data partners through OHDSIs standard recruitment process, which includes protocol publication on OHDSIs GitHub, an announcement in OHDSIs research forum, presentation at the weekly OHDSI all-hands-on meeting and direct requests to data holders.

Table \@ref(tab:data-sources) lists the 5 already committed data sources; these sources encompass a large variety of practice types and populations. 
For each data source, we report a brief description and size of the population it represents.
All data sources will receive institutional review board approval or exemption for their participation before executing the study.

```{r data-sources, echo=FALSE, warning=FALSE}
data_sources <- readr::read_delim(col_names = TRUE, delim = ";", trim_ws = TRUE, file = "
  Data source ; Population ; Patients ; History ; Data capture process and short description
  IBM MarketScan Commercial Claims and Encounters (CCAE) ; Commercially insured, < 65 years ; 142M ; 2000 -- ; Adjudicated health insurance claims (e.g. inpatient, outpatient, and outpatient pharmacy)  from large employers and health plans who provide private healthcare coverage to employees, their spouses and dependents.
  IBM MarketScan Medicare Supplemental Database (MDCR)  ; Commercially insured, 65$+$ years ; 10M ; 2000 -- ; Adjudicated health insurance claims of retirees with primary or Medicare supplemental coverage through privately insured fee-for-service, point-of-service or capitated health plans.
  IBM MarketScan Multi-State Medicaid Database (MDCD) ; Medicaid enrollees, racially diverse ; 26M ; 2006 -- ; Adjudicated health insurance claims for Medicaid enrollees from multiple states and includes hospital discharge diagnoses, outpatient diagnoses and procedures, and outpatient pharmacy claims.
  Optum Clinformatics Data Mart (Optum) ; Commercially or Medicare insured ; 85M ; 2000 -- ; Inpatient and outpatient healthcare insurance claims.
  Optum Electronic Health Records (OptumEHR) ; US, general ; 93M ; 2006 -- ; Clinical information, prescriptions, lab results, vital signs, body measurements, diagnoses and procedures derived from clinical notes using natural language processing. 
")
tab <- kable(data_sources, booktabs = TRUE, linesep = "",
      caption = "Committed  data sources and the populations they cover.") %>% 
  kable_styling(bootstrap_options = "striped", latex_options = "striped") %>%
  pack_rows("Administrative claims", 1, 4, latex_align = "c", indent = FALSE) %>%
  pack_rows("Electronic health records (EHRs)", 5, 5, latex_align = "c", indent = FALSE)
if (knitr::is_latex_output()) {
  tab %>%
    column_spec(1, width = "10em") %>%
    column_spec(2, width = "10em") %>%
    column_spec(5, width = "25em") %>%
    kable_styling(font_size = latex_table_font_size)
} else {
  tab
}
```


# Strengths and Limitations

## Strengths



## Limitations

- We are unable to test our models in a clinical setting
- In a clinical setting predictors may be self reported and these may differ from our covariate definitions
- The electronic health record databases may be missing care episodes for patients due to care outside the respective health systems.

# Protection of Human Subjects

This study does not involve human subjects research.
The project does, however, use de-identified human data collected during routine healthcare provision.
All data partners executing the study within their data sources will have received institutional review board (IRB) approval or waiver for participation in accordance to their institutional governance prior to execution (see Table \@ref(tab:irb)).
This study executes across a federated and distributed data network, where analysis code is sent to participating data partners and only aggregate summary statistics are returned, with no sharing of patient-level data between organizations.

```{r irb, echo=FALSE,warning=FALSE}
data_sources <- readr::read_delim(col_names = TRUE, delim = "&", trim_ws = TRUE, file = "
Data source & Statement
IBM MarketScan Commercial Claims and Encounters (CCAE) & New England Institutional Review Board and was determined to be exempt from broad IRB approval, as this research project did not involve human subject research.
IBM MarketScan Medicare Supplemental Database (MDCR)  & New England Institutional Review Board and was determined to be exempt from broad IRB approval, as this research project did not involve human subject research.
IBM MarketScan Multi-State Medicaid Database (MDCD) & New England Institutional Review Board and was determined to be exempt from broad IRB approval, as this research project did not involve human subject research.
Optum Clinformatics Data Mart (Optum) & New England Institutional Review Board and was determined to be exempt from broad IRB approval, as this research project did not involve human subject research.
Optum Electronic Health Records (OptumEHR) & New England Institutional Review Board and was determined to be exempt from broad IRB approval, as this research project did not involve human subject research.
")
tab <- kable(data_sources, booktabs = TRUE, linesep = "",
      caption = "IRB approval or waiver statement from partners.") %>% 
  kable_styling(bootstrap_options = "striped", latex_options = "striped")
if (knitr::is_latex_output()) {
  tab %>%
    column_spec(1, width = "15em") %>%
    column_spec(2, width = "40em") %>%
    kable_styling(font_size = latex_table_font_size)
} else {
  tab
}
```

# Plans for Disseminating and Communicating Study Results

Open science aims to make scientific research, including its data process and software, and its dissemination, through publication and presentation, accessible to all levels of an inquiring society, amateur or professional [@Woelfle2011-ss] and is a governing principle of this study.
Open science delivers reproducible, transparent and reliable evidence.
All aspects of this study (except private patient data) will be open and we will actively encourage other interested researchers, clinicians and patients to participate.
This differs fundamentally from traditional studies that rarely open their analytic tools or share all result artifacts, and inform the community about hard-to-verify conclusions at completion.

## Transparent and re-usable research tools

We will publicly register this protocol and announce its availability for feedback from stakeholders, the OHDSI community and within clinical professional societies.
This protocol will link to open source code for all steps to generate and evaluate prediction models, figures and tables.
Such transparency is possible because we will construct our studies on top of the OHDSI toolstack of open source software tools that are community developed and rigorously tested [@Schuemie2020-wx].
We will publicly host the source code at (https://github.com/ohdsi-studies/FeatureSelectionComparison), allowing public contribution and review, and free re-use for anyone’s future research.

## Scientific meetings and publications

We will deliver multiple presentations at scientific venues and will also prepare multiple scientific publications for clinical, informatics and statistical journals.

## General public

We believe in sharing our findings that will guide clinical care with the general public.
We will use social-media (Twitter) to facilitate this.


